---
title: "Final Report_Yihan Zhong"
author: "Yihan Zhong"
date: "2026-05-04"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    latex_engine: xelatex
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(haven)
library(dplyr)
library(ggplot2)
library(knitr)
library(tidyr)
library(scales)
```

# Introduction

## Research Question

Is smoking prevalence associated with education level among U.S. adults?
Specifically, does the probability of being a current smoker decrease
as educational attainment increases?

## Scientific Plausibility and Context

Tobacco use remains one of the leading preventable causes of death
globally, responsible for an estimated 6.4 million deaths annually
[@Latvala2022]. In the United States, cigarette smoking is not uniformly
distributed across the population: epidemiological evidence consistently
demonstrates an inverse association between educational attainment and
smoking prevalence, with lower-educated individuals being more likely to
smoke and less likely to successfully quit [@Garrett2019].

This disparity has widened over time. Analysis of National Health
Interview Survey data spanning 1966 to 2018 shows that smoking
disparities by education have increased across birth cohorts, with
college-educated individuals exhibiting the lowest initiation
probabilities and highest cessation rates across all age groups and
genders [@Cao2023].

Several mechanisms may explain this pattern. Higher education is
associated with greater health literacy, enabling individuals to better
understand the health risks of tobacco use and navigate cessation
resources. Additionally, individuals with lower educational attainment
face greater socioeconomic stressors and are more frequently targeted
by tobacco industry marketing [@Cao2023]. Evidence from a pooled
analysis of 18 twin cohorts further supports a causal interpretation:
within pairs of twins discordant for both education and smoking,
the less-educated co-twin was consistently more likely to be a current
smoker, even after controlling for shared genetic and family background
[@Latvala2022].

Based on this evidence, we hypothesize that smoking prevalence will
decrease monotonically with increasing educational attainment among
U.S. adults in the 2024 BRFSS dataset.

# Material & Methods

## Data

The data used in this report come from the 2024 Behavioral Risk Factor
Surveillance System (BRFSS), an annual, cross-sectional telephone survey
conducted by the U.S. Centers for Disease Control and Prevention (CDC).
The BRFSS collects state-level data on health behaviors, chronic
conditions, and preventive health practices among U.S. adults aged
18 years and older. The 2024 dataset was downloaded from the CDC BRFSS
website in XPT format.

Two variables were selected for this analysis:

- **`_RFSMOK3`**: A derived variable indicating current smoking status
  (1 = current smoker, 2 = not a current smoker). Respondents with
  missing or ambiguous values were excluded.
- **`EDUCA`**: Self-reported highest level of education completed,
  coded on a 6-point ordinal scale. Respondents with missing values
  or codes outside 1–6 were excluded.

For analysis, `EDUCA` was recoded into three ordered groups:
**Low** (codes 1–3: less than high school), **Medium** (code 4: high
school graduate or GED), and **High** (codes 5–6: some college or
college graduate). Smoking status was converted to a binary variable
(`current_smoker`: 1 = yes, 0 = no).

```{r data-load, results='hide'}
setwd("C:/Users/HUAWEI/Desktop/MPH 2026 spring/6270/cp03/LLCP2024XPT")
brfss <- read_xpt("LLCP2024.XPT")

brfss_clean <- brfss %>%
  dplyr::select(`_RFSMOK3`, EDUCA) %>%
  dplyr::filter(
    !is.na(`_RFSMOK3`),
    !is.na(EDUCA),
    EDUCA %in% 1:6,
    `_RFSMOK3` %in% c(1, 2)
  ) %>%
  mutate(
    current_smoker = ifelse(`_RFSMOK3` == 2, 1, 0),
    edu_group = case_when(
      EDUCA %in% 1:3 ~ "Low",
      EDUCA == 4     ~ "Medium",
      EDUCA %in% 5:6 ~ "High"
    ),
    edu_group = factor(edu_group, levels = c("Low", "Medium", "High"))
  )
```

```{r variable-table, echo=FALSE}
variable_info <- data.frame(
  `Variable Name` = c("_RFSMOK3", "EDUCA", "current_smoker", "edu_group"),
  Description = c(
    "Current smoking status (derived)",
    "Highest education level completed",
    "Binary smoker indicator (1=Yes, 0=No)",
    "Grouped education (Low/Medium/High)"
  ),
  `Variable Type` = c("Categorical", "Categorical (ordinal)", "Binary", "Categorical (ordinal)"),
  `R Class` = c("numeric", "numeric", "numeric", "factor"),
  check.names = FALSE
)
kable(variable_info, caption = "Overview of Selected Variables")
```

## Statistical Analyses

The association between education level and current smoking status was
examined using two complementary approaches.

**Chi-square test of independence.** A Pearson chi-square test was used
to assess whether smoking prevalence differed significantly across
education groups. The assumptions of the chi-square test require that
all expected cell frequencies are >= 5, which was verified prior to
analysis. The effect size was quantified using Cramér's V, where values
of 0.1, 0.3, and 0.5 indicate small, medium, and large associations,
respectively.

**Logistic regression.** To quantify the association between education
and smoking while accounting for the ordinal nature of the exposure,
a binary logistic regression model was fitted with `current_smoker` as
the outcome and `edu_group` as the predictor, using the High education
group as the reference category. Odds ratios (OR) and 95% confidence
intervals (CI) are reported. A trend test was additionally conducted
by treating education as a numeric ordinal variable (Low = 1,
Medium = 2, High = 3) to assess whether smoking odds changed
monotonically across groups.

All analyses were conducted in R. The significance threshold was set
at $\alpha$ = 0.05.

## Code

All code and data documentation required to replicate this report are
available at: **https://github.com/muxiao1019-bot/VTPEH6270-CP06**

# Results

## Visualization

```{r prevalence-plot, fig.height=4, fig.width=6, fig.cap="Smoking prevalence decreases monotonically with higher educational attainment (2024 BRFSS)."}
smoke_by_edu <- brfss_clean %>%
  group_by(edu_group) %>%
  summarise(
    n = n(),
    smokers = sum(current_smoker),
    smoking_prev = mean(current_smoker),
    .groups = "drop"
  )

ggplot(smoke_by_edu, aes(x = edu_group, y = smoking_prev, fill = edu_group)) +
  geom_col(width = 0.6, show.legend = FALSE) +
  geom_text(aes(label = paste0(round(smoking_prev * 100, 1), "%")),
            vjust = -0.5, size = 3.5) +
  scale_fill_manual(values = c("Low" = "#d73027",
                                "Medium" = "#fc8d59",
                                "High" = "#4575b4")) +
  scale_y_continuous(labels = percent_format(), limits = c(0, 0.35)) +
  theme_minimal(base_size = 12) +
  labs(
    title = "Smoking Prevalence by Education Level",
    x = "Education Level",
    y = "Smoking Prevalence"
  )
```

## Statistical Analyses

### Chi-square Test

```{r chisq-assumption, results='hide', echo=FALSE}
# Contingency table
cont_table <- table(brfss_clean$edu_group, brfss_clean$current_smoker)
kable(cont_table, caption = "Contingency Table: Education Group vs Smoking Status")

# Expected frequencies
chisq_result <- chisq.test(cont_table)
expected <- chisq_result$expected
kable(round(expected, 1), caption = "Expected Cell Frequencies (all must be >= 5)")
```

All expected cell frequencies were well above 5, confirming that the
chi-square test assumption was met.

```{r chisq-test, echo=FALSE}
# Chi-square test result
chisq_result

# Cramer's V effect size
n_total <- sum(cont_table)
cramers_v <- sqrt(chisq_result$statistic /
                    (n_total * (min(nrow(cont_table),
                                   ncol(cont_table)) - 1)))
cat("Cramer's V:", round(cramers_v, 4), "\n")
```

### Logistic Regression

```{r logistic-regression, echo=FALSE}
# Logistic regression: High as reference group
brfss_clean$edu_group <- relevel(brfss_clean$edu_group, ref = "High")
model_cat <- glm(current_smoker ~ edu_group,
                 data = brfss_clean,
                 family = binomial)

# Odds ratios and 95% CI
or_table <- exp(cbind(OR = coef(model_cat),
                      confint(model_cat)))
kable(round(or_table, 3),
      caption = "Odds Ratios from Logistic Regression (Reference: High education)")
```

```{r trend-test, echo=FALSE}
# Trend test: education as ordinal numeric variable
brfss_clean <- brfss_clean %>%
  mutate(edu_numeric = case_when(
    edu_group == "Low"    ~ 1,
    edu_group == "Medium" ~ 2,
    edu_group == "High"   ~ 3
  ))

model_trend <- glm(current_smoker ~ edu_numeric,
                   data = brfss_clean,
                   family = binomial)

trend_or <- exp(cbind(OR = coef(model_trend),
                      confint(model_trend)))
kable(round(trend_or, 3),
      caption = "Trend Test: OR per Unit Increase in Education Level")
```

```{r or-plot, fig.height=3.5, fig.width=6, fig.cap="Odds ratios for current smoking by education group relative to High education (logistic regression)."}
# Forest plot of ORs
or_df <- data.frame(
  Group = c("Low vs High", "Medium vs High"),
  OR    = exp(coef(model_cat)[2:3]),
  Lower = exp(confint(model_cat)[2:3, 1]),
  Upper = exp(confint(model_cat)[2:3, 2])
)

ggplot(or_df, aes(x = OR, y = Group)) +
  geom_point(size = 3, color = "#d73027") +
  geom_errorbarh(aes(xmin = Lower, xmax = Upper),
                 height = 0.2, color = "#d73027") +
  geom_vline(xintercept = 1, linetype = "dashed", color = "gray40") +
  scale_x_log10() +
  theme_minimal(base_size = 12) +
  labs(
    title = "Odds Ratios for Current Smoking by Education Group",
    x = "Odds Ratio (log scale)",
    y = NULL
  )

```

The chi-square test revealed a statistically significant association
between education level and smoking status ($\chi^2$= 9223.4, df = 2,
p < 0.001). Cramér's V = 0.148, indicating a small-to-moderate
effect size.

Logistic regression showed that compared to individuals with high
educational attainment, those with low education had significantly
higher odds of being a current smoker
(OR = 3.34, 95% CI: 3.23–3.45),
as did those with medium education
(OR = 2.24, 95% CI: 2.20–2.29).
The trend test confirmed a significant monotonic decrease in smoking
odds with increasing education (OR per unit increase = 0.51, p < 0.001).


# Conclusions

This analysis of the 2024 BRFSS dataset provides strong evidence of an
inverse association between educational attainment and current smoking
prevalence among U.S. adults. Both the chi-square test and logistic
regression confirmed that individuals with lower education were
significantly more likely to be current smokers, with a clear
monotonic trend across the three education groups. These findings are
consistent with the existing literature documenting persistent and
widening educational disparities in tobacco use in the United States
[@Garrett2019; @Cao2023].

The observed pattern aligns with proposed mechanisms linking education
to smoking behavior, including greater health literacy and access to
cessation resources among more educated individuals [@Latvala2022].
The large sample size of the BRFSS — consistent with simulations
conducted in Checkpoint 04 — ensured high statistical power to detect
even modest effect sizes, as reflected by the highly significant
chi-square result. However, the relatively small Cramér's V value
suggests that while the association is statistically robust, education
alone explains a limited proportion of the variance in smoking behavior,
and other socioeconomic and contextual factors likely contribute
to the observed disparities.

From a public health perspective, these findings underscore the
importance of targeting tobacco control efforts toward lower-educated
populations, who bear a `d`isproportionate burden of smoking-related
disease and mortality.

# AI Use Disclosure Statement

This document was prepared with assistance from Claude (Anthropic).
The tool was used to generate the initial R Markdown structure and 
develop the statistical analysis code, all code and interpretations were reviewed and verified by the author.

# References
